---
title: "ps7"
author: "Hannah Olson-Williams"
date: "12/5/2021"
output: word_document
---

### QUESTION 2 PART A 

Intrinsic rate of natural increase for the russian female population of 2000 

Is nma the same as B/N?! 
nma = rate of bearing female children between ages a and a+5 
B/N = female births / female population 

preston p114: NRR = 1/l0 * sum(nFx*nLx)
where nFx = nBx/nWx 
```{r}
setwd("H:/soci756")
d= read.csv("ps7.csv")
d = d[is.na(d$age) == FALSE,] #remove na row 

#remove commas and reformat nlx values 
d$nLx = as.numeric(gsub(",","",d$nLx))

#calculate the nMa values: rate of bearing female children between ages a and a+5
d$nma = d$B/ d$N
d$lm = d$nma* d$nLa

#subset data for maternal ages only 
dmat = d[d$age >10 & d$age <=45,]
r0 = log(sum(dmat$nLa*dmat$nma, na.rm = TRUE))/27 

NRR = sum(dmat$nLa*dmat$nma, na.rm = TRUE)


for (a in 15:45) {
yrn = function(rn){
  exp(-rn*(a+2.5))*d$nLa[d$age == a]*d$nma[d$age == a]
}}

r1 = r0 + ((yrn(r0)-1)/27)
r2 = r1 + ((yrn(r1)-1)/27)
r3 = r2 + ((yrn(r2)-1)/27)
r4 = r3 + ((yrn(r3)-1)/27)
r5 = r4 + ((yrn(r4)-1)/27)
```
The NRR of the population is `r NRR`. 
After three iterations, we obtain `r r3` for the intrinsic growth rate of Russia in 2000. 


### QUESTION 2 PART B
The "real" mean length of generation in the Russian population of 2000:

NRR = e^(r*t) -> ln(NRR) = r*t -> t = ln(NRR)/ r 
```{r}
t = log(NRR) / r3 

t
```
The number of years required for the population to grow (at the intrinsic growth rate) by the factor equal to the 2000 next reproduction rate is `r t`. 

### QUESTION 2 PART C
Intrinsic age distribution of the Russian female population in 2000 
```{r}
d$bdenom = exp(-r3*(d$age+2.5))*(d$nLa/100000)
b = 1/sum(d$bdenom)


d$cstable = b*d$bdenom 
d$cactual = d$N / sum(d$N)

#print entire dataframe 
d
```

### QUESTION 2 PART D 
```{r}
library(ggplot2)

#make a new dataset to play nicely with ggplot 
dplot = data.frame(age = rep(d$age, 2), c = c(d$cstable, d$cactual), group = c(rep("Intrinsic", 21), rep("Actual", 21)))

ggplot(dplot) + geom_point(aes(x = age, y = c, shape = group, color = group), size = 3) +
  xlab("Age a") + 
  ylab("5Ca") +
  ggtitle("Actual versus Intrinsic Age Distributions")+
  theme_bw() +
  theme(legend.title = element_blank())

```
The difference between the age distribution in Russia in 2000 and the age distribution of the stable equivalent population is that the stable equivalent population grows until expected age of death at around age 80. On the other hand, the actual age distribution in Russia steadily declines after age 50 as as age increases. The actual age distribution follows a less strict pattern and seemingly includes greater variation in birth and death rates.  
Given that the intrinsic growth rate r3 that I calculated in part a has approximately 20 times greater magnitude than CRNI = -0.006, it makes sense that the actual and intrinsic age distributions are substantially different. Additionally, as stated on Preston page 148, populations in Europe and North America are expected to have large discrepancies between the intrinsic growth rate and the CRNI since large changes in fertility and mortality rates have "destabilized" these populations over time.    

### QUESTION 2 PART E 
Intrinsic death rate 
```{r}
idr = b-r3
idr
```
The intrinsic death rate is `r idr`. 





### QUESTION 3 

a) Simulate a population with only two birth cohorts of 10,000 people each 

```{r}

#this is the beginning of what will be a lifetable style table with each row representing a different age  
c1 = data.frame(age = 1:100)
c2 = data.frame(age = 1:100)

#this is the beginning of what will be an individual level table with each id representing an unique member of the population 
p1 = data.frame(id = 1:10000)
p2 = data.frame(id = 1:10000)

```

b) Assign frailty values that are normally distributed 
I tried to use a gamma distribution to avoid negative frailty values. 
```{r}
#need one gamma dist value per inndividual
for (i in 1:nrow(p1)) { 
  p1$frailty[i] = rgamma(1,2)
  p2$frailty[i] = rgamma(1,2)
}

#sort frailty values such that the largest frailty values have the smallest indices 
p1 = p1[order(p1$frailty, decreasing = TRUE),]
p2 = p2[order(p2$frailty, decreasing = TRUE),]



```

c) Generate a mortality function that ignores mortality before age 35 and assumes gompertz mortality after age 35. 
Set mortality in cohort2 to be 30% higher than mortality in cohort1 

Gompertz: u(age) = alpha * e^(Beta)*age 

```{r}
alpha = 0.002 #these values were selected from Alexander's plot 
Beta = 0.05 


#death rates are 1.3 times greater in cohort 2 than in cohort 1 
#this is the gompertz function mu(x) 
#when age is <35, i do not calculate a mortality rate
c1$mort = as.numeric(ifelse(c1$age>=35, alpha*exp(Beta*c1$age), ""))
c2$mort = as.numeric(c1$mort)*1.3

#initialize new column to store the number of survivors 
c1$nalive = NA  
c2$nalive = NA

#when age is <35, assume that no one dies since no mortality -> survivors = l0 
c1$nalive[1:34] = 10000 
c2$nalive[1:34] = 10000

#use mortality value to calculate number of survivors at each age >=35 
for (i in 35:length(c1$age)){
  c1$nalive[i] = c1$nalive[i-1] - (c1$nalive[i-1]*c1$mort[i])
  c2$nalive[i] = c2$nalive[i-1] - (c2$nalive[i-1]*c2$mort[i])
}

#calculate frailty for each age  
#initialize frail columns 
c1$frail = rep(0,length(c1$age)) 
c2$frail = rep(0,length(c2$age))

#take average of frailty values when age <35 (this is necessary to create initial conditions for loop below)
c1$frail[35] = mean(p1$frailty[1:round(c1$nalive[35], 0)])
c2$frail[35] = mean(p2$frailty[1:round(c2$nalive[35], 0)])

#find mean frailty of individuals still alive at each age (note that the most frail die first/youngest)
for (i in 36:length(c1$age)){
 c1$frail[i] = mean(p1$frailty[round(c1$nalive[i-1],0):round(c1$nalive[i], 0)])
 c2$frail[i] = mean(p1$frailty[round(c2$nalive[i-1],0):round(c2$nalive[i], 0)])
}




pi0 = 0.95 #chosen based on values used in vaupel and yashin 
c1$pi = (pi0*c1$nalive)/ (pi0*c1$nalive + (1-pi0)*c2$nalive) #this is eq3 from vaupel and yashin
c1$meanmort = c1$pi*c1$mort + (1-c1$pi)*c2$mort #eq4 from vaupel and yashin 
c1$meanfrail = c1$pi*c1$frail + (1-c1$pi)*c2$frail #I assume that eq3 and eq4 hold for frailty values as well as mortality 

```






### PLOTS OF MORTALITY RATES AND FRAILTY BY AGE FOR EACH COHORT AND AS A WHOLE 
```{r}

#make one dataset for easier plotting 
# (i probably should have been working from a single dataset this entire time.....)
pc1 = c1[c1$age > 34,] #remove ages under 35 since they don't contain any information 
pc2 = c2[c2$age > 34,]
pdata = data.frame(age = rep(pc1$age,3), mort = c(pc1$mort, pc2$mort, pc1$meanmort), frail = c(pc1$frail, pc2$frail, pc1$meanfrail), cohort = c(rep("Cohort 1", 66), rep("Cohort 2",66), rep("Overall",66)))



#plot for mortality rates versus age
#note that I did not log transform these values.... 
ggplot(pdata) + geom_point(aes(x=age, y=mort, shape = cohort, color = cohort), size = 3)+
  xlab("Age") + 
  ylab("Mortality") +
  ggtitle("Mortality by Age")+
  theme_bw() +
  theme(legend.title = element_blank())

#plot for frailty versus age 
ggplot(pdata) + geom_point(aes(x=age, y=frail, shape = cohort, color = cohort), size=3)+
  xlab("Age") + 
  ylab("Frailty") +
  ggtitle("Frailty by Age")+
  theme_bw() +
  theme(legend.title = element_blank())
```



